The Host Galaxies of Rapidly Evolving Transients in the Dark Energy Survey

### 2020-02-27 RET_Hosts.tex

Recently, inspection of high-cadence, all-sky survey data sets have revealed yet more exotic transients that are less easy to explain with conventional models. \citet{Drout2014} revealed a class of rapidly evolving transients (RETs; also termed `Fast Blue Optical Transients' - FBOTs or `Fast Evolving Luminous Transients' - FELTs) in the Pan-STARRS survey (PS1). \citet{Pursiainen2018} expanded the known number of RETs to beyond 80 with their sample from the Dark Energy Survey (DES), spanning a redshift range of $\sim 0$ to $>1$. RETs typically rise to peak brightness within 7-10 days, and decline to 10\% of their peak brightness within 30 days, much faster than typical SNe. The photometric measurements of the PS1 and DES RETs seem to be well described by simple expanding blackbodies, although a handful show declining photospheric radii from the outset. Due to the rapid nature of their lightcurves and location at high-redshift, spectral coverage is sparse and signal-to-noise ratio (SNR) is low, such that there has not yet been a conclusive detection of absorption or emission features from the transients and thus the physical mechanism responsible for their rapid evolution remains unexplained.

@@

We derive our sample from the 106 RETs discovered in the 5-year DES-SN transient survey. This number expands upon the 72 of \citetalias{Pursiainen2018}. The first reason for the increased sample size is the use of the 5th year of DES-SN, where \citetalias{Pursiainen2018} were only able to make use of the first four years. The second reason is an update to the sample selection technique. By imposing the \citetalias{Pursiainen2018} selection criteria on season 5, the sample is increased to 96 objects.

@@

To assess the distinguishing photometric properties of RETs, we take the sample of 72 from the \citetalias{Pursiainen2018} method along with spectroscopically confirmed SNe of types Ia, Ibc, II, and SLSNe from DES that pass the following criteria:
\begin{enumerate}
\item The transient was only detected in one DES-SN observing season.
\item Maximum observed brightness in both $g$- and $r$-bands was brighter than 24 mag (in the eight `shallow' DES-SN fields) or 25 mag (in the two `deep' fields).
\item $g$- and $r$-band observations used for the colour had to be taken within 2 days in observer frame.
\end{enumerate}

Of objects passing these cuts, the RETs cluster at shorter timescales and bluer colours than other SNe, even in the observer frame. However, using photometric data points directly has several problems that can be improved. For one, measuring peak colour is problematic: DES-SN did not always observe $g$- and $r$-bands on the same or even consecutive nights, thus making it impossible to measure the peak colour in a number of cases. Measuring rise times of 10-15 days is difficult to perform with a one week cadence when it has to be done without fitting a light curve model, hence the rise time values are spread over a wide range. To negate this issue, we interpolate the light curves using the method of Gaussian processes (GP). The interpolated light curves have a 0.5 day cadence and every epoch has a flux value, and an associated uncertainty, for each band. Using this technique, SNe Ia and RETs populate two distinct regions of $g-r$-$t_{\mathrm{rise; r}}$ parameter space (Fig. \ref{fig:selection}) and are thus easier to separate. We define a region in this parameter space which minimises the contamination of non-RETs (purity) while maximising the total fraction of RETs (completeness). The resulting limits are $-1.5 < g-r < 0.6 $ and $t_{\mathrm{rise; r}} <16.5$, and the parameter space can be seen in Fig. \ref{fig:selection}.

We process all $\sim 30,000$ DES-SN transient candidates with GP. In order to reduce the contamination from active galactic nuclei (AGN), we use a basic convolutional neural network (CNN) classifier which we train on spectrosopically confirmed objects, and use to separate the sample into two subtypes: AGN-like and SNe-like. The classifier returns SNe-like objects with an accuracy of 0.992 on the test set; the remaining AGN are likely removed by the final manual vetting.
The SNe passing the CNN classifier are subjected to the LC quality cuts, resulting in 2259 objects, of which 939 lie inside the colour and rise-time region which we defined as RETs. These objects are subject to a further set of cuts. We impose a cut based on a fit of the LC with the \texttt{PSNID} software (\citet{Sako2008}). We use thresholds of \texttt{FITPROB}$<0.91$ and \texttt{PBAYES}$<0.82$ to remove highly-probably SNe Ia, which removes 46 objects from the RET parameter space. In order to further remove longer-lived SNe, the decline time to half of the peak brightness must be $<24$ days. This removes 347 SNe, resulting in 546 objects remaining inside the parameter space. The final 564 transients have been visually inspected, with the majority rejected for being spurious detections, obvious multi-season variability that was not picked up by the CNN, or showing evidence for a longer timescale decline.

Using the above method recovers N of the 96 rets found using the \citetalias{Pursiainen2018} technique, and adds a further M.
We refer to the resulting sample as DES RETs. Of the 106 objects in the sample, 97 have a host galaxy detected in deep host galaxy photometry of \citet{Wiseman2020}, of which 49 have a host galaxy spectroscopic redshift. A further three have redshifts obtained from narrow lines observed in spectra of the transients themselves. We do not consider these three objects for the analysis, since we are unable to separate transient and host contributions to the spectra. The observational properties of the 49 hosts with redshift are displayed in Table \ref{tab:obs}.

@@

\subsubsection{RETs \label{subsubsec:compare_rets}}
Since the DES sample of RETs is by far the largest discovered to date, there is no other large sample of RETs with which to compare host galaxy properties. \citet{Drout2014} present host galaxies of 10 RETs discovered in the Pan-STARRS survey, with measurements of stellar masses and SFRs. To this we add the low-redshift transient AT2018cow with host galaxy photometric measurements from \citet{Perley2019} and metallicity (\citealt{Pettini2004} O3N2) from \citet{Morokuma-Matsui}.

@@

\subsection{Comparison with AT2018cow \label{subsec:disc_cow}}

The nearby transient AT2018cow has drawn many comparisons to the cosmological RETs from DES and PS1 \citep[e.g.][]{Perley2019,Margutti2019,Mohan2020,Fox2020} due to its rapid evolution and blue colour.

>>>>

Recently, inspection of high-cadence, all-sky survey data sets have revealed yet more exotic transients that are less easy to explain with conventional models. \citet{Drout2014} revealed a class of rapidly evolving transients (RETs; also termed `Fast Blue Optical Transients' - FBOTs or `Fast Evolving Luminous Transients' - FELTs) in the Pan-STARRS survey (PS1). \citet{Pursiainen2018} expanded the known number of RETs to beyond 80 with their sample from the Dark Energy Survey (DES), spanning a redshift range of $\sim 0$ to $>1$. RETs typically rise to peak brightness within 7-10 days \textcolor{green}{(Would prefer "less than 10d" in DES at least we can't measure 7-10 with certainity)}, and decline to 10\% of their peak brightness within 30 days, much faster than typical SNe. The photometric measurements of the PS1 and DES RETs seem to be well described by simple expanding blackbodies, although a handful show declining photospheric radii from the outset. Due to the rapid nature of their lightcurves and location at high-redshift, spectral coverage is sparse and signal-to-noise ratio (SNR) is low, such that there has not yet been a conclusive detection of absorption or emission features from the transients and thus the physical mechanism responsible for their rapid evolution remains unexplained.

@@

We derive our sample from the 106 RETs discovered in the 5-year DES-SN transient survey. This number expands upon the 72 of \citetalias{Pursiainen2018}. The first reason for the increased sample size is the use of the 5th year of DES-SN, where \citetalias{Pursiainen2018} were only able to make use of the first four years. The second reason is an update to the sample selection technique. By imposing the \citetalias{Pursiainen2018} selection criteria on season 5, the sample is increased to 92 objects.

@@

To assess the distinguishing photometric properties of RETs, we take the sample of 72 \textcolor{green}{(92)} from the \citetalias{Pursiainen2018} method along with spectroscopically confirmed SNe of types Ia, Ibc, II, and SLSNe from DES that pass the following criteria:
\begin{enumerate}
\item The transient was only detected in one DES-SN observing season.
\item Maximum observed brightness in both $g$- and $r$-bands was brighter than 24 mag (in the eight `shallow' DES-SN fields) or 25 mag (in the two `deep' fields).
\item $g$- and $r$-band observations used for the colour had to be taken within 2 days in observer frame.
\end{enumerate}

Of objects passing these cuts, the RETs cluster at shorter timescales and bluer colours than other SNe, even in the observer frame. However, using photometric data points directly has several problems that can be improved. For one, measuring peak colour is problematic: DES-SN did not always observe $g$- and $r$-bands on the same or even consecutive nights, thus making it impossible to measure the peak colour in a number of cases. Measuring rise times of 10-15 days is difficult to perform with a one week cadence when it has to be done without fitting a light curve model, hence the rise time values are spread over a wide range. To negate this issue, we interpolate the light curves using the method of Gaussian processes (GP). \textcolor{green}{(To negate this issue, we use the interpolated GP LCs presented in Pursiainen et al. 2019)} The interpolated light curves have a 0.5 day cadence and every epoch has a flux value, and an associated uncertainty, for each band. Using this technique, SNe Ia and RETs populate two distinct regions of $g-r$-$t_{\mathrm{rise; r}}$ parameter space (Fig. \ref{fig:selection}) and are thus easier to separate. We define a region in this parameter space which minimises the contamination of non-RETs (purity) while maximising the total fraction of RETs (completeness). The resulting limits are $-1.5 < g-r < 0.6 $ and $t_{\mathrm{rise; r}} <16.5$, and the parameter space can be seen in Fig. \ref{fig:selection}.

We process all $\sim 30,000$ DES-SN transient candidates with GP. In order to reduce the contamination from active galactic nuclei (AGN), we use a basic convolutional neural network (CNN) classifier which we train on spectrosopically confirmed objects, and use to separate the sample into two subtypes: AGN-like and SNe-like. The classifier returns SNe-like objects with an accuracy of 0.992 on the test set; the remaining AGN are likely removed by the final manual vetting. \textcolor{green}{(I'll find the list of AGN and find out where they are from.)}
The SNe passing the CNN classifier are subjected to the LC quality cuts, resulting in 2259 objects, of which 939 lie inside the colour and rise-time region which we defined as RETs. These objects are subject to a further set of cuts. We impose a cut based on a fit of the LC with the \texttt{PSNID} software (\citet{Sako2008}). We use thresholds of \texttt{FITPROB}$<0.91$ and \texttt{PBAYES}$<0.82$ to remove highly-probably SNe Ia, which removes 46 objects from the RET parameter space. In order to further remove longer-lived SNe, the decline time to half of the peak brightness must be $<24$ days. This removes 347 SNe, resulting in 546 objects remaining inside the parameter space. The final 564 transients have been visually inspected, with the majority rejected for being spurious detections, obvious multi-season variability that was not picked up by the CNN, or showing evidence for a longer timescale decline.

Using the above method recovers 87 of the 92 RETs found using the \citetalias{Pursiainen2018} technique, and adds a further 14. \textcolor{green}{The five were not recovered as their GP lightcurves were fainter than the limits given above in either g or r band.}
We refer to the resulting sample as DES RETs. Of the 106 objects in the sample, 97 have a host galaxy detected in deep host galaxy photometry of \citet{Wiseman2020}, of which 49 have a host galaxy spectroscopic redshift. A further three have redshifts obtained from narrow lines observed in spectra of the transients themselves. We do not consider these three objects for the analysis, since we are unable to separate transient and host contributions to the spectra. The observational properties of the 49 hosts with redshift are displayed in Table \ref{tab:obs}.

@@

\subsubsection{RETs \label{subsubsec:compare_rets}}
Since the DES sample of RETs is by far the largest discovered to date, there is no other large sample of RETs with which to compare host galaxy properties. \citet{Drout2014} present host galaxies of 10 RETs discovered in the Pan-STARRS survey, with measurements of stellar masses and SFRs. To this we add the low-redshift transient AT2018cow with host galaxy photometric measurements from \citet{Perley2019} and metallicity (\citealt{Pettini2004} O3N2) from \citet{Morokuma-Matsui}, and SN2018gep with data from \citet{Ho2019}.

@@

\subsection{Comparison with local RETs \label{subsec:disc_lowz}}

The nearby transient AT2018cow has drawn many comparisons to the cosmological RETs from DES and PS1 \citep[e.g.][]{Perley2019,Margutti2019,Mohan2020,Fox2020} due to its rapid evolution and blue colour. AT2018cow displayed a contracting photosphere as well as evidence for central-engine power alongside an unusual spectrum that showed similarities to broad-lined SNe Ic (SN Ic-bl) at early stages \citep[e.g.][]{Xu2018,Izzo2018} , developing to something entirely different later on \citep{Perley2019} with hints of similarities to interacting SNe Ibn \citep{Fox2020}. There have been several suggestions that AT2018cow is indeed an analogue of the high-z RETs. The host galaxy of AT2018cow appears to be moderately star forming and lies very close to the centre of the SFMS (Figs. \ref{fig:sfms_sfr},\ref{fig:sfms_ssfr}), along with many of the DES RET hosts. However, the host lies significantly above the fiducial MZR in Fig. \ref{fig:mzr}, suggesting that it has an abnormally high metallicity for its stellar mass. This is in contrast to the DES RET hosts, which are systematically less enriched for a given stellar mass.

Other local rapid transients include SN2018gep \citep{Ho2019}, a spectroscopically classified SN Ic-bl with a rapid rise. The host of SN2018gep appears more similar to the DES RET sample, lying in the same $M_*$-SFR and $M*$-sSFR plane, as well as lying below the MZR. The authors' conclusion that SN2018gep is related to a shock-breakout of a massive, stripped-envelope star is similar to that posited in Section \ref{subsec:disc_origin}. SN2018kzr \citep{McBrien2019} is one of the most rapidly declining transients ever discovered, with spectral signatures similar to SNe Ic. While host galaxy properties are not derived, the authors of that paper refer to narrow emission from the host galaxy, along with an apparently small, blue, star-forming host.
